---
title: 'Part 1: Data'
author: "Anna, Riley"
date: "2/7/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, warning = FALSE,
                      fig.width = 5, fig.height = 3.5, 
                      fig.align = 'center')

library(tidyverse)
library(broom)
library(tidymodels)
```

# Overview of the Dataset

```{r}
Batter_Metrics
```

The dataset will use are information on baseball players of ___team collected from 2021, in which individual baseball players are the observational unit. Corresponding independent categorical variables we will consider are (1) handedness, whether the player bats with his right or left arm; (2) position, (3) ; (4) . The independent quantitative variables we will consider are (1) ; (2) ; (3) ; (4) ; (5) ; (6) ; (7) ; (8) ; (9) ; (10) ; (11) ; (12) .
    
    - what is population
    - sample representation of population
    - missing observations?
    
    
# Summary Statistics 
```{r, echo=FALSE}
#bb_data <- read_csv("")
#bb_model <- lm( ~ , data = bb_data) 
```
- indicate what the statistics say about your data. Try to examine binary or other relationships as well as describing individual distributions.



# Graphical Analysis

- Graphical displays with adequate explanation / interpretation (These should effectively summarize your data and point out any interesting features. 

- quantitative vs quantitative - categories in colors
- no linear models

# Discussion
- discuss the limitations of describing a larger population.